Categories of Optics

### 2018-06-18 paper.tex
Work on Traversals section

% Categories
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Cat}{\mathbf{Cat}}
\newcommand{\Prof}{\mathbf{Prof}}
\newcommand{\Core}{\mathbf{Core}}
\newcommand{\MonCat}{\mathbf{MonCat}}
\newcommand{\LaxMonCat}{\mathbf{LaxMonCat}}
\newcommand{\SymmMonCat}{\mathbf{SymmMonCat}}
\newcommand{\Tele}{\mathbf{Tele}}

\newcommand{\Act}{\mathbf{Act}}
\newcommand{\Optic}{\mathbf{Optic}}
\newcommand{\Lawful}{\mathbf{Lawful}}
\newcommand{\SemiOptic}{\mathbf{SemiOptic}}
\newcommand{\Lens}{\mathbf{Lens}}
\newcommand{\Prism}{\mathbf{Prism}}
\newcommand{\Setter}{\mathbf{Setter}}
\newcommand{\Hask}{\mathbf{Hask}}
\newcommand{\Endo}{\mathbf{Endo}}
\newcommand{\Strong}{\mathbf{Strong}}
\newcommand{\Tamb}{\mathbf{Tamb}}
\newcommand{\Point}{\mathbf{Point}}
\newcommand{\CoPoint}{\mathbf{CoPoint}}
\newcommand{\Traversable}{\mathbf{Traversable}}

@@

  Now let $\rep{l}{r}$ be a representative for $p$, so $rl = \id_S$ and $\langle lr \rangle = \langle \psi A\rangle$. There therefore exists a finite chain of relations $lr = u_1 \leftrightsquigarrow \dots \leftrightsquigarrow u_n = \psi A$.

Suppose the first relation faces rightward, so there exists a $k$ and $\phi$ with $r;l = \phi_A;k$ and $u_2 = k;\phi_A$. Define $l' = l;\phi_A$ and $r' = k;r$. Then:
\begin{align*}
\rep{l'}{r'} 
&= \rep{l ; \phi_A}{k ; r} \\
&= \rep{l}{\phi_A ; k ; r} \\
&= \rep{l}{r ; l; r} \\
&= \rep{l}{r}
\end{align*}
This new representative satisfies
\begin{align*}
l';r' &= l;\phi_A;k;r = l;r;l;r = \id_S \\
r';l' &= k;r;l;\phi_A = k;\phi_A;k;\phi_A = u_2^2
\end{align*}

@@

The general pattern is as follows. Once we choose a particular monoidal action $\M \to [\C, \C]$, we find an isomorphism between the set of optics $(S, S') \hto (A, A')$ and a set $\mathbb{O}((S, S'), (A, A'))$ that is easier to describe. Ideally, we then also find a simplified description $\mathbb{D}(S, A)$ for the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)\] appearing in the definition of a lawful optic. We can then ``read off'' what conditions on an element of $\mathbb{O}((S, S), (A, A))$ are needed to ensure that the corresponding element of $\Optic_\M(S, A)$ is lawful. With no offence intended, we will call these more specific conditions the `ad-hoc' laws.

@@

@@

\begin{definition}
  Suppose $\C$ has finite products, and let $\C$ act on itself via the bifunctor $\times : \C \times \C \to \C$. The \emph{category of lenses} $\Lens$ is the category of lawful optics with respect to this action: $\Lens = \Lawful_\times$.
\end{definition}

@@

If we ask more of our category $\C$, we can upgrade this result to show that a lens $S \hto A$ implies the existence of a complement $\C$ with $S \cong C \times A$. This is a generalised from the argument given in~\cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

\begin{proposition}
  Suppose $\C$ has pullbacks and that there is a morphism $x : 1 \to A$. If $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
  Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
  \[
    \begin{tikzcd}
      S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
      & C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
      & 1\ar[r, "x", swap] & A
    \end{tikzcd}
  \]
  which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
  \begin{align*}
    \fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
                                &= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
                                &= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
                                &= \id_S && \text{(by $\fget\fput$)}
                                            \intertext{and}
                                            [j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
                                &= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
                                &= [j\fput (i \times A), \pi_2] && \\
                                &= [jij\fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
                                &= [jij \pi_1 (i \times A), \pi_2] && \\
                                &= [jiji \pi_1, \pi_2] && \\
                                &= [\pi_1, \pi_2] && \\
                                &= \id_{C \times A}
  \end{align*}

@@

\begin{definition}
  Suppose $\C$ has finite coproducts, and let $\C$ act on itself via the bifunctor $\sqcup : \C \times \C \to \C$. The \emph{category of prisms} is the category of lawful optics with respect to this action: $\Prism = \Lawful_\sqcup$.
\end{definition}

@@

\begin{definition}
  Supposed $\C$ is cartesian closed and let $\Strong(\C)$ act on $\C$ by evaluation. The \emph{category of setters} $\Setter_\C$ is the category of optics for this action.
\end{definition}

@@

@@

A folklore variant of a prism is the \emph{affine traversal}, which allows access to a target that may or may not be present. These are described in~\cite{AffineTraversalPost}, and it is suggested they correspond to pairs of maps $\C(S, A \sqcup S') \times \C(S\times A', S')$. This is not precisely correct, but very close:

@@

\begin{remark}
  It is important here that the morphisms in $\mathsf{Aff}$ are only those that arise from pairs of morphisms $P \to P'$ and $Q \to Q'$, although in principle there are other natural transformations between the corresponding functors $(P, Q) \cdot -$ and $(P', Q') \cdot -$.
\end{remark}

@@

This set includes into $\C(S, A \sqcup S') \times \C(S\times A', S')$, but the two are not equal.

@@

\subsection{Traversals}
\todo{todo}

>>>>

% Categories
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Cat}{\mathbf{Cat}}
%\newcommand{\Hask}{\mathbf{Hask}}
\newcommand{\Prof}{\mathbf{Prof}}
\newcommand{\Core}{\mathbf{Core}}
\newcommand{\MonCat}{\mathbf{MonCat}}
\newcommand{\LaxMonCat}{\mathbf{LaxMonCat}}
\newcommand{\SymmMonCat}{\mathbf{SymmMonCat}}
\newcommand{\Tele}{\mathbf{Tele}}
\newcommand{\Tamb}{\mathbf{Tamb}}

\newcommand{\Endo}{\mathbf{Endo}}
\newcommand{\Strong}{\mathbf{Strong}}
\newcommand{\Point}{\mathbf{Point}}
\newcommand{\CoPoint}{\mathbf{CoPoint}}
\newcommand{\Traversable}{\mathbf{Traversable}}

\newcommand{\Act}{\mathbf{Act}}
\newcommand{\Optic}{\mathbf{Optic}}
\newcommand{\Lawful}{\mathbf{Lawful}}
%\newcommand{\SemiOptic}{\mathbf{SemiOptic}}
\newcommand{\Lens}{\mathbf{Lens}}
\newcommand{\Prism}{\mathbf{Prism}}
\newcommand{\Setter}{\mathbf{Setter}}
\newcommand{\Traversal}{\mathbf{Traversal}}

@@

  Now let $\rep{l}{r}$ be a representative for $p$, so $rl = \id_S$ and $\langle lr \rangle = \langle \psi A\rangle$. There therefore exists a finite chain of relations $lr = u_1 \leftrightsquigarrow \dots \leftrightsquigarrow u_n = \psi A$. Suppose the first relation faces rightward, so there exists a $k$ and $\phi$ with $r;l = \phi_A;k$ and $u_2 = k;\phi_A$. Define $l' = l;\phi_A$ and $r' = k;r$. Then:
\begin{align*}
\rep{l'}{r'} 
&= \rep{l ; \phi_A}{k ; r} \\
&= \rep{l}{\phi_A ; k ; r} \\
&= \rep{l}{r ; l; r} \\
&= \rep{l}{r}
\end{align*}
This new representative satisfies
\begin{align*}
l';r' &= l;\phi_A;k;r = l;r;l;r = \id_S \\
r';l' &= k;r;l;\phi_A = k;\phi_A;k;\phi_A = u_2^2
\end{align*}

@@

The general pattern is as follows. Once we choose a particular monoidal action $\M \to [\C, \C]$, we find an isomorphism between the set of optics $(S, S') \hto (A, A')$ and a set $\mathbb{O}((S, S'), (A, A'))$ that is easier to describe. We follow \cite{ProfunctorOptics} (and others) in calling elements of $\mathbb{O}$ \emph{concrete optics}. There is no canonical choice for this set; our primary goal is to find a way to eliminate the coend so that we no longer have to deal with equivalence classes of morphisms.

Ideally, we then also find a simplified description $\mathbb{D}(S, A)$ for the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)\] appearing in the definition of a lawful optic. We can then ``read off'' what conditions on are needed on a concrete optic to ensure that the corresponding element of $\Optic_\M(S, A)$ is lawful. With no offence intended, we will call these more conditions the `ad-hoc' laws.

@@

We begin with the most common kinds of optic before moving to more exotic possibilities.

@@

\begin{definition}
  Suppose $\C$ has finite products, and let $\C$ act on itself via the bifunctor $\times : \C \times \C \to \C$. The \emph{category of lenses} $\Lens$ is the category of lawful optics with respect to this action.
\end{definition}

@@

If we ask more of our category $\C$, we can upgrade this result to show that a lens $S \hto A$ implies the existence of a complement $\C$ with $S \cong C \times A$. This doesn't appear to follow purely from the ad-hoc lens laws---an argument using a definition of lawfulness based on constant complements.

\begin{proposition}[{Generalisation of \cite[Corollary 13]{AlgebrasAndUpdateStrategies}}]
  Suppose $\C$ has pullbacks and that there is a morphism $x : 1 \to A$. If $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
  Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
  \[
    \begin{tikzcd}
      S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
      & C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
      & 1\ar[r, "x", swap] & A
    \end{tikzcd}
  \]
  which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
  \begin{align*}
    \fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
                                &= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
                                &= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
                                &= \id_S && \text{(by $\fget\fput$)}
                                            \intertext{and}
                                            [j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
                                &= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
                                &= [j\fput (i \times A), \pi_2] && \\
                                &= [jij\fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
                                &= [jij \pi_1 (i \times A), \pi_2] && \\
                                &= [jiji \pi_1, \pi_2] && \\
                                &= [\pi_1, \pi_2] && \\
                                &= \id_{C \times A}
  \end{align*}

@@

\begin{definition}
  Suppose $\C$ has finite coproducts, and let $\C$ act on itself via the bifunctor $\sqcup : \C \times \C \to \C$. The \emph{category of prisms} is the category of lawful optics with respect to $\sqcup$.
\end{definition}

@@

\begin{definition}
  Supposed $\C$ is cartesian closed and let $\Strong(\C)$ act on $\C$ by evaluation. The \emph{category of setters} $\Setter_\C$ is the category of lawful optics for this action.
\end{definition}

@@

\subsection{Traversals}
Let $\C = \Set$. \todo{define applicative/traversable functors}

\begin{definition}
The category $\Traversal_\C$ of traversals in $\C$ is the category of lawful optics for the action of $\Traversable_\C$ on $\C$ given by evaluation.
\end{definition}

(Yes, the names $\Traversal$ vs\@. $\Traversable$ are confusing.)

It is known that traversable functors correspond to coalgebras for a particular parameterised comonad. See~\cite{AlgebrasForParameterisedMonads} for the relevant definitions of parameterised monads and algebras.

\begin{proposition}[{\cite[Theorem 4.10, Proposition 5.4]{SecondOrderFunctionals}}]
Traversable structures on a functor $T : \Set \to \Set$ correspond to parameterised coalgebra structures
\begin{align*}
t_{A, B} : TA \to UR^*_{A, B}(T B)
\end{align*} 
where $UR^*_{X,Y}$ is the parameterised comonad
\begin{align*}
UR^*_{X, Y} Z = \Sigma_{n\in \bN} X^n \times \homC(Y^n,Z)
\end{align*}
Moreover, this correspondence forms an isomorphism of categories between $\Traversable$ and the Eilenberg-Moore category of coalgebras for $UR^*_{-, -}$, which we denote $\C_T$.
\end{proposition}

\begin{remark}
In~\cite[Section 2.3]{ProfunctorOptics}, it is claimed that a traversal $S \hto A$ exhibits an isomorphism \[S \cong \Sigma_n A^n \times \homC(A^n,S).\] This appears to be a misreading of~\cite[Proposition 5.4]{SecondOrderFunctionals}.
\end{remark}

The language of coalgebras now allows us to characterise traversals.

\begin{lemma}
  For any objects $A, B \in \Set$ and traversable functor $F$, \[\C(FA, B) \cong \Traversable(F, \Sigma_n (-)^n \times \homC(A^n,B))\]
naturally in $B$ and $F$. In other words, the functor \[(B \mapsto \Sigma_n (-)^n \times \homC(A^n,B)) : \C \to \Traversable\] is right adjoint to the evaluation-at-$A$ functor $-A : \Traversable \to \C$.
\end{lemma}
\begin{proof}
By~\cite[Proposition 6]{AlgebrasForParameterisedMonads}, there is a parameterised adjunction $L_T \dashv R_T$, where
\begin{align*}
L_T : \C^\op \times \C_T &\to \C \\
(X, (F, f)) &\mapsto FX \\
R_T : \C \times \C &\to \C_T \\
(Y, Z) &\mapsto (UR^*_{-, Y} Z, \varepsilon) \\
\end{align*}
where $\varepsilon$ is the counit of $UR^*_{-, -}$. Evaluating these with the fixed parameter $A$, we get an ordinary adjunction 
\begin{align*}
L_T(A) : \C_T &\to \C \\
(F, f) &\mapsto FA \\
R_T(A) : \C &\to \C_T \\
Z &\mapsto (UR^*_{-, A} Z, \varepsilon) \\
\end{align*}
But this is exactly the adjunction we were trying to show.
\end{proof}

We then reach the same characterisation of traversals as~\cite{ProfunctorOptics}:
\begin{align*}
\Optic_\Traversable((S, S'), (A, A')) &= \int^{F \in \Traversable} \C(S, F A) \times \C(F A', S') \\
&\cong \int^{F \in \Traversable} \C(S, F A) \times \Traversable(F, \Sigma_n (-)^n \times \homC(A'^n,S')) \\
&\cong \C(S, \Sigma_n A^n \times \homC(A'^n,S')) \\
&=: \mathbb{O}((S, S'), (A, A'))
\end{align*}

\todo{laws}

\begin{remark}
\todo{todo?} A careful analysis would be needed to describe traversals in some other category, in particular, the description of $UR^*$ does not make sense if $\C$ is not locally cartesian closed.
\end{remark}

\begin{remark}
Traversable functors can be described as a particular class of polynomial functors known as finitary containers. There may be something generalisable here with other classes of polynomial functors.
\end{remark}

@@

A folklore variant of a prism is the \emph{affine traversal}, which allows access to a target that may or may not be present.

@@

\begin{remark}
  It is important here that the morphisms in $\mathsf{Aff}$ are only those that arise from pairs of morphisms $P \to P'$ and $Q \to Q'$, although in principle there may be other natural transformations between the corresponding functors $(P, Q) \cdot -$ and $(P', Q') \cdot -$.
\end{remark}

@@

We note that this contradicts the folklore description of affine traversals as pairs of maps $\C(S, A \sqcup S') \times \C(S\times A', S')$, see~\cite{AffineTraversalPost}. That set is in fact ``too large'': the provided map $S \times A' \to S$ must be defined even on elements of $S$ whose image under the first map $S \to A \sqcup S'$ lies in the right-hand component, which is not the case in our description.

@@
